---
title: Source Data for Figure 1
author: "Tommaso Biancalani and Jeff Gore"
output:
  pdf_document: default
  html_notebook: default
---

Clear environment and set working directories.

```{r}
rm(list=ls())
wd1 <- './../exp1/'
wd2 <- './../exp2/'
```

# Linear fit for lethalities $\delta$ in survival curves in Fig. 1-A and Fig. S1

Load file lists of each experimental condition.

```{r}
paA = list.files(path=wd2, pattern="paA[[:alpha:]]")
paB = list.files(path=wd2, pattern="paB[[:alpha:]]")
paC = list.files(path=wd2, pattern="paC[[:alpha:]]")
smA = list.files(path=wd2, pattern="smA[[:alpha:]]")
smB = list.files(path=wd2, pattern="smB[[:alpha:]]")
smC = list.files(path=wd2, pattern="smC[[:alpha:]]")
seA = list.files(path=wd2, pattern="seA[[:alpha:]]")
seB = list.files(path=wd2, pattern="seB[[:alpha:]]")
seC = list.files(path=wd2, pattern="seC[[:alpha:]]")
```

Load data from CSV files.

```{r}
read_csv_list <- function(wd, csv_files) {
  # Read list of CSV files and return list of corresponding dataframes
  dfs <- list()
  for (csv_file in csv_files) {
    csv_file = paste(wd, csv_file, sep = "")
    df <- read.csv(csv_file)
    dfs <- c(dfs, list(df))
  }
  return(dfs)
}

PaA_dfs = read_csv_list(wd2, paA)
PaB_dfs = read_csv_list(wd2, paB)
PaC_dfs = read_csv_list(wd2, paC)

SmA_dfs = read_csv_list(wd2, smA)
SmB_dfs = read_csv_list(wd2, smB)
SmC_dfs = read_csv_list(wd2, smC)

SeA_dfs = read_csv_list(wd2, seA)
SeB_dfs = read_csv_list(wd2, seB)
SeC_dfs = read_csv_list(wd2, seC)
```

Normalize survival curves to get fraction of worms surviving on y-axis.

```{r}
normalize_survival_curves <- function (dfs) {
  # Take list of survival curves and return normalized list of survival curves
  
  norm_dfs <- list()
  for (df in dfs) {
    n_worms = df[[1, 2]]
    df[[2]] = df[[2]] / n_worms
    norm_dfs <- c(norm_dfs, list(df))
  }  
  return(norm_dfs)
}

PaA_dfs = normalize_survival_curves(PaA_dfs)
PaB_dfs = normalize_survival_curves(PaB_dfs)
PaC_dfs = normalize_survival_curves(PaC_dfs)

SmA_dfs = normalize_survival_curves(SmA_dfs)
SmB_dfs = normalize_survival_curves(SmB_dfs)
SmC_dfs = normalize_survival_curves(SmC_dfs)

SeA_dfs = normalize_survival_curves(SeA_dfs)
SeB_dfs = normalize_survival_curves(SeB_dfs)
SeC_dfs = normalize_survival_curves(SeC_dfs)

```

Display mean survival curve to detect invasion time (vertical bar), which is used to determine the fitting region.

```{r}
display_mean_surv_curve <- function (dfs, title_text, num_pts) {
  # Display mean surv. curve averaged iover list of dataframes `dfs`
  # Set figure title to `title_text`
  # Draw vertical line on plot to separate last `num_pts`
  # Return `num_pts`
  
  ## Get mean survival curve
  w_rows <- list()
  for (df in dfs) {
    w_rows <- as.double(c(w_rows, df[[2]]))
  }
  row_matrix <- matrix(w_rows, nrow = length(dfs), byrow = TRUE)
  mean_sc <- colMeans(row_matrix)
  
  ## Plot
  times <- dfs[[1]][[1]]
  xlab <- 'Time (hr)'
  ylab <- 'Fraction of worms surviving'
  plot(times, mean_sc, log = 'y', type = 'b', main = title_text, xlab = xlab, ylab = ylab)
  
  ## Draw vertical line
  threshold_time <- rev(times)[[num_pts]]
  abline(v=threshold_time)
  
  return(num_pts)
}
```

For *P. aeruginosa*:

```{r}
PaA_npts <- display_mean_surv_curve(PaA_dfs, 'Pa 48h', 3)
PaB_npts <- display_mean_surv_curve(PaB_dfs, 'Pa 24h', 4)
PaC_npts <- display_mean_surv_curve(PaC_dfs, 'Pa 4h', 4)
```

For *S. marcescens*:

```{r}
SmA_npts <- display_mean_surv_curve(SmA_dfs, 'Sm 48h', 3)
SmB_npts <- display_mean_surv_curve(SmB_dfs, 'Sm 24h', 3)
SmC_npts <- display_mean_surv_curve(SmC_dfs, 'Sm 4h', 3)
```

For *S. enterica*:

```{r}
SeA_npts <- display_mean_surv_curve(SeA_dfs, 'Se 48h', 3)
SeB_npts <- display_mean_surv_curve(SeB_dfs, 'Se 24h', 3)
SeC_npts <- display_mean_surv_curve(SeC_dfs, 'Se 4h', 3)
```

Fit lethality to each survival curve using a linear model. The fitting region is determined by the invasion times from the mean survival curves.

```{r}
find_models <- function (dfs, num_pts) {
  # Take last `num_pts` from each df from `dfs, 
  # Compute log of y and perfom linear fit.
  # return fitted models
  models <- list()
  for (df in dfs) {
    
    ## Convert to semi-log
    last_pts = tail(df, num_pts)
    t <- last_pts[[1]]
    w <- log(last_pts[[2]])
    
    ## Purge non-valid values
    mask <- is.finite(w)
    w <- w[mask]  
    t <- t[mask]
    
    ## Linear fit
    model <- lm(w ~ t)
    models <- c(models, list(model))
  }
  return(models)
}

PaA_models = find_models(PaA_dfs, PaA_npts)
PaB_models = find_models(PaB_dfs, PaB_npts)
PaC_models = find_models(PaC_dfs, PaC_npts)

SmA_models = find_models(SmA_dfs, SmA_npts)
SmB_models = find_models(SmB_dfs, SmB_npts)
SmC_models = find_models(SmC_dfs, SmC_npts)

SeA_models = find_models(SeA_dfs, SeA_npts)
SeB_models = find_models(SeB_dfs, SeB_npts)
SeC_models = find_models(SeC_dfs, SeC_npts)
```

Display linear fits.

```{r}
display_surv_fit <- function (dfs, models, title_text) {
  # Display overlayed surv. curves. from list of dataframes `dfs`
  # Set figure title to `title_text`
  # Overlay linear fit from `models`.
  
  ## Main plot
  df <- dfs[[1]]
  title <- title_text
  xlab <- 'Time (hr)'
  ylab <- 'Fraction of worms surviving'
  t <- df[[1]]
  w <- log(df[[2]])
  options <- list(
    t, w, col=1, type='p', main=title, xlab=xlab, ylab=ylab
    )
  do.call(plot, options)
  
  ## Overlay secondary plots
  for (i in 2:length(dfs)) {
    df <- dfs[[i]]
      t <- df[[1]]
    w <- log(df[[2]])
    points(t, w, col=i, type='p')
  }
  
  ## Overlay linear fit
  for (i in 1:length(models)) {
    model <- models[[i]]
    t <- df[[1]]
    w <- log(df[[2]])
    col <- i
    abline(model, col=i, lwd=.6, lt=2)
  }
}

display_surv_fit(PaA_dfs, PaA_models, "Pa 48h")
display_surv_fit(PaB_dfs, PaB_models, "Pa 24h")
display_surv_fit(PaC_dfs, PaC_models, "Pa 4h")

display_surv_fit(SmA_dfs, SmA_models, "Sm 48h")
display_surv_fit(SmB_dfs, SmB_models, "Sm 24h")
display_surv_fit(SmC_dfs, SmC_models, "Sm 4h")

display_surv_fit(SeA_dfs, SeA_models, "Se 48h")
display_surv_fit(SeB_dfs, SeB_models, "Se 24h")
display_surv_fit(SeC_dfs, SeC_models, "Se 4h")
```

Compute mean lethalities and their standard errors.

```{r}
get_delta_w_err <- function (models) {
  deltas <- list() 
  for (model in models) {
    delta <- model$coefficients[[2]]
    deltas <- c(deltas, delta)
  }
  deltas <- unlist(deltas)  # cast to vector
  delta <- -mean(deltas)
  delta <- round(delta, digits = 3)
  err <- sqrt(var(deltas)/length(deltas))
  err <- round(err, digits = 3)
  return (c(delta, err))
}

PaA_delta  <- get_delta_w_err(PaA_models)
PaB_delta  <- get_delta_w_err(PaB_models)
PaC_delta  <- get_delta_w_err(PaC_models)

SmA_delta  <- get_delta_w_err(SmA_models)
SmB_delta  <- get_delta_w_err(SmB_models)
SmC_delta  <- get_delta_w_err(SmC_models)

SeA_delta  <- get_delta_w_err(SeA_models)
SeB_delta  <- get_delta_w_err(SeB_models)
SeC_delta  <- get_delta_w_err(SeC_models)

Pa_delta_df= data.frame(c("average lethality", "standard error"), PaA_delta, PaB_delta, PaC_delta, row.names = 1)
colnames(Pa_delta_df) <- c("Pa 48 hr", "Pa 24 hr", "Pa 4 hr")
Pa_delta_df

Sm_delta_df= data.frame(c("average lethality", "standard error"), SmA_delta, SmB_delta, SmC_delta, row.names = 1)
colnames(Sm_delta_df) <- c("Sm 48 hr", "Sm 24 hr", "Sm 4 hr")
Sm_delta_df

Se_delta_df= data.frame(c("average lethality", "standard error"), SeA_delta, SeB_delta, SeC_delta, row.names = 1)
colnames(Se_delta_df) <- c("Se 48 hr", "Se 24 hr", "Se 4 hr")
Se_delta_df
```

# Linear fit for lethalities $\delta$ in survival curves in Fig. 1-B and Fig. S2

Find file list corresponding to technical replica

```{r}
pa_fls = list.files(path=wd1, pattern="paW[[:alnum:]]")
sm_fls = list.files(path=wd1, pattern="smW[[:alnum:]]")
se_fls = list.files(path=wd1, pattern="seW[[:alnum:]]")
ph_fls = list.files(path=wd1, pattern="phW[[:alnum:]]")
```

Compute normalized survival curves from CSV files.

```{r}
pa_dfs = read_csv_list(wd1, pa_fls)
pa_dfs = normalize_survival_curves(pa_dfs)

sm_dfs = read_csv_list(wd1, sm_fls)
sm_dfs = normalize_survival_curves(sm_dfs)

se_dfs = read_csv_list(wd1, se_fls)
se_dfs = normalize_survival_curves(se_dfs)

ph_dfs = read_csv_list(wd1, ph_fls)
ph_dfs = normalize_survival_curves(ph_dfs)
```

Display survival groups and determine threshold of exponential phase to be used for fitting.

```{r}
pa_npts <-  display_mean_surv_curve(pa_dfs, "Pa", 4)
sm_npts <- display_mean_surv_curve(sm_dfs, "Sm", 5)
se_npts <- display_mean_surv_curve(se_dfs, "Se", 4)
ph_npts <- display_mean_surv_curve(ph_dfs, "Ph", 4)
```

Fit lethality with linear model using thresholds previously found.

```{r}
pa_models = find_models(pa_dfs, pa_npts)
sm_models = find_models(sm_dfs, sm_npts)
se_models = find_models(se_dfs, se_npts)
ph_models = find_models(ph_dfs, ph_npts)
```

Check that fits are performed correctly.

```{r}
display_surv_fit(pa_dfs, pa_models, "Pa")
display_surv_fit(sm_dfs, sm_models, "Sm")
display_surv_fit(se_dfs, se_models, "Se")
display_surv_fit(ph_dfs, ph_models, "Ph")
```

Compute average lethalities with s.e.m

```{r}
pa_delta  <- get_delta_w_err(pa_models)
sm_delta  <- get_delta_w_err(sm_models)
se_delta  <- get_delta_w_err(se_models)
ph_delta  <- get_delta_w_err(ph_models)

res <- data.frame(c("average lethality", "standard error"), pa_delta, sm_delta, se_delta, ph_delta, row.names = 1)
colnames(res) <- c("Pa", "Sm", "Se", "Ph")
res

```
